---
author: "Notes from: AMF"
title: "SEER - Breast cancer: invasive vs in situ"
date: "11/8/2023"
output: pdf_document
---


## Data source

Institute NC. Surveillance, Epidemiology, and End Results (SEER 8, and 13,) Program Populations (1975-2020). (www.seer.cancer.gov/popdata), National Cancer Institute, DCCPS, Surveillance Research Program, released May 2023. February 2022S ed.: National Cancer Institute, 2022.

### Datasets available
#### SEER8
SEER8 Research plus - IDC and DCIS by, age, year, ER status, from 1979 to 2020:
DCIS:https://raw.githubusercontent.com/filhoalm/Breast_cancer/main/dataCheck/dcis.csv"
IDC: https://raw.githubusercontent.com/filhoalm/Breast_cancer/main/dataCheck/idc.csv

#### SEER13
SEER13 Research plus - Malignant breast cancer by race and ethnicity, year, age, ER status, HER status from 1992 to 2018: https://github.com/filhoalm/Breast_cancer/blob/main/forecasting/data/breast_er_her_11072023.csv

SEER13 Research plus - Malignant breast cancer by year, age, ER status, HER status, PR status from 1992 to 2018: 
https://github.com/filhoalm/Breast_cancer/blob/main/forecasting/data/breast_er_her_pr_1182023.csv


#### Morphology

###### DCIS

'8201/2: Cribriform carcinoma in situ','8500/2: Intraductal carcinoma, noninfiltrating, NOS','8501/2: Comedocarcinoma, noninfiltrating','8503/2: Noninfiltrating intraductal papillary adenocarcinoma','8507/2: Intraductal micropapillary carcinoma','8523/2: Intraductal with other types of carcinoma in situ' AND {Extent of Disease.ER Status Recode Breast Cancer (1990+)} = 'Positive','Negative','Borderline/Unknown','Recode not available'

###### IDC

'8500/3: Infiltrating duct carcinoma, NOS','8523/3: Infiltrating duct mixed with other types of carcinoma' AND {Extent of Disease.ER Status Recode Breast Cancer (1990+)} = 'Positive','Negative','Borderline/Unknown','Recode not available'



##### Temporal trends of female breast cancer by subtype and ER status, using SEER8, 1979 - 2020.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(Rcan)
library(ggplot2)
library(dplyr)
library(ggplot2)
library(gridExtra)
```

```{r load_data}
df1 <- read.csv("https://raw.githubusercontent.com/filhoalm/Breast_cancer/main/dataCheck/dcis.csv")
names(df1) <- c("year", "age", "er", "asr", "count", "py", "std")
df1$cancer <- "DCIS"

df2 <- read.csv("https://raw.githubusercontent.com/filhoalm/Breast_cancer/main/dataCheck/idc.csv")
names(df2) <- c("year", "age", "er", "asr", "count", "py", "std")
df2$cancer <- "IDC"

```


```{r load_data2}

# Define a function to perform repetitive operations
prepare_data <- function(df1, df2, age_index_range, start_year) {
  df <- bind_rows(df1, df2) %>%
    mutate(year = year + 1974,
           er_label = case_when(
             er == 0 ~ "positive",
             er == 1 ~ "negative",
             er == 2 ~ "bord/unkw",
             TRUE ~ "not available"
           ),
           age_group = cut(age, breaks = seq(29, 85, 5), labels = paste(seq(30, 80, 5), "-", seq(34, 84, 5))),
           age_index = as.integer(age_group) + 6) %>%
    filter(age >= 30 & age <= 84 & year >= 1975, age_index %in% age_index_range)
  
  df_selected <- select(df, c(1, 5, 6, 8, 9, 11))
  
  result <- csu_asr(df_selected, "age_index", "count", "py",
                    group_by = c("cancer", "er_label", "year"),
                    var_age_group = c("er_label", "year"))
  
  subset(result, year >= start_year)
}

# Compute ASR for each age group
df6 <- prepare_data(df1, df2, 7:10, 1992) # 30-49 yo
df7 <- prepare_data(df1, df2, 11:17, 1992) # 50-84 yo


```


```{r load_data3}

# # Define a function to create the plot
create_asr_plot <- function(df) {
  ggplot(df, aes(x = year, y = asr, color = er_label, shape = cancer)) +
    geom_point() +
    geom_line() +
    labs(x = "Year", y = "ASR", color = "ER Label", shape = "Cancer") +
    theme_minimal()
}
```


```{r load_data4}

# Create the plots
plot2 <- create_asr_plot(df6) + labs(title = "30-49 yo") # ASR for the age group 30-49
plot3 <- create_asr_plot(df7) + labs(title = "50-84yo") # ASR for the age group 50-84+

# Then use grid.arrange to put them side by side
grid.arrange(plot2, plot3, ncol = 2)
 
 



```




##### Is Ductal Carcinoma in Situ a Precursor or Low-Risk Variant of Invasive Breast Cancer? A Population Perspective


```{r}

library(ggplot2)
library(gridExtra)
library(dplyr)

#MyPath <- "C:/Users/filhoam/Desktop/Breast/Round2/Figure/"


# Function for reading and processing data
process_df <- function(file_path, index) {
  df = read.csv(file_path)
  df$year <- df$Year.of.diagnosis + 1974
  names(df) <- c("year_num", "er", "asr", "cases", "py", "year")
  df$er_label <- factor(df$er, labels = c("ER+", "ER-", "Borderline/Unk", "Not available"))
  refs <- df$asr[df$year == 1990]
  df <- df %>% 
    group_by(er) %>% 
    mutate(rr = round((asr/refs[er + 1]), 2)) %>% 
    ungroup()
  df$index <- index
  df$site <- "DCIS"
  return(df)
}

# File paths
file_paths <- c('file_path1' = 'https://raw.githubusercontent.com/filhoalm/EpicanVerse/main/dcis_er_30_49.csv',
                'file_path2' = 'https://raw.githubusercontent.com/filhoalm/EpicanVerse/main/dcis_er_50_84.csv',
                'file_path3' = 'https://raw.githubusercontent.com/filhoalm/EpicanVerse/main/dcis_er.csv'
)

# Indices
indices <- c('30_49', '50_84', '30_84')

# Process dataframes
dfs <- mapply(process_df, file_path = file_paths, index = indices, SIMPLIFY = F)

# Combine dataframes
dcis <- do.call(rbind, dfs)

# IDC

# Function for reading and processing data
process_df <- function(file_path, index) {
  df = read.csv(file_path)
  df$year <- df$Year.of.diagnosis + 1974
  names(df) <- c("year_num", "er", "asr", "cases", "py", "year")
  df$er_label <- factor(df$er, labels = c("ER+", "ER-", "Borderline/Unk", "Not available"))
  refs <- df$asr[df$year == 1990]
  df <- df %>% 
    group_by(er) %>% 
    mutate(rr = round((asr/refs[er + 1]), 2)) %>% 
    ungroup()
  df$index <- index
  df$site <- "IDC"
  return(df)
}

# File paths
file_paths <- c('file_path1' = 'https://raw.githubusercontent.com/filhoalm/EpicanVerse/main/idc_er_30_49.csv',
                'file_path2' = 'https://raw.githubusercontent.com/filhoalm/EpicanVerse/main/idc_er_50_84.csv',
                'file_path3' = 'https://raw.githubusercontent.com/filhoalm/EpicanVerse/main/idc_er.csv'
)

# Indices
indices <- c('30_49', '50_84', '30_84')

# Process dataframes
dfs <- mapply(process_df, file_path = file_paths, index = indices, SIMPLIFY = F)

# Combine dataframes
idc <- do.call(rbind, dfs)


data <- rbind(dcis, idc)

generate_plot <- function(data, title, ylim_val) {
  title = paste0(title)
  plot <- ggplot(data, aes(x=year, y= rr, group=er_label)) +
    geom_point(aes(shape=er_label, color=er_label)) +
    scale_color_manual(values=c('#374e55','#df8f44','darkgray'))+ 
    ylim (0, ylim_val) +
    labs(x="Year", y="Incidence relative to 1990") +
    geom_hline(yintercept=1, linetype="dashed", size=0.3) +
    theme_bw() +
    theme(legend.position="none") +
    ggtitle(title)+
    theme(plot.title = element_text(size = 5),
          axis.text.x = element_text(size = 5),
          axis.text.y = element_text(size = 5),
          axis.title.x = element_text(size = 5),
          axis.title.y = element_text(size = 5))
  return(plot)
}

# Define the subset of your data for each index
indices <- list("30_84", "30_49", "50_84")

#
a <- subset(data, data$site=="IDC" & data$index=="30_84" &data$er_label!="Not available" & data$year >= 1990)
b <- subset(data, data$site=="IDC" & data$index=="30_49" &data$er_label!="Not available" & data$year >= 1990)
c <- subset(data, data$site=="IDC" & data$index=="50_84" &data$er_label!="Not available" & data$year >= 1990)

a1 <- subset(data, data$site=="DCIS" & data$index=="30_84" &data$er_label!="Not available" & data$year >= 1990)
b1 <- subset(data, data$site=="DCIS" & data$index=="30_49" &data$er_label!="Not available" & data$year >= 1990)
c1 <- subset(data, data$site=="DCIS" & data$index=="50_84" &data$er_label!="Not available" & data$year >= 1990)


plot30_84 <- generate_plot(a, '30 - 84', 5) + 
  ggtitle(paste0('IDC, ', gsub('_', ' - ', '30 - 84')))
assign(paste0('plot', '30 - 84'), plot)

plot30_49 <- generate_plot(b, '30 - 49', 5) + 
  ggtitle(paste0('IDC, ', gsub('_', ' - ', '30 - 49')))
assign(paste0('plot', '30 - 49'), plot)

plot50_84 <- generate_plot(c, '50 - 84', 5) + 
  ggtitle(paste0('IDC, ', gsub('_', ' - ', '50 - 84')))
assign(paste0('plot', '50 - 84'), plot)

#

plotA30_84 <- generate_plot(a1, '30 - 84', 20) + 
  ggtitle(paste0('DCIS, ', gsub('_', ' - ', '30 - 84')))
assign(paste0('plot', '30 - 84'), plot)

plotA30_49 <- generate_plot(b1, '30 - 49', 20) + 
  ggtitle(paste0('DCIS, ', gsub('_', ' - ', '30 - 49')))
assign(paste0('plot', '30 - 49'), plot)

plotA50_84 <- generate_plot(c1, '50 - 84', 20) + 
  ggtitle(paste0('DCIS, ', gsub('_', ' - ', '50 - 84')))
assign(paste0('plot', '50 - 84'), plot)

grid.arrange(plot30_84, plot30_49, plot50_84,
             plotA30_84, plotA30_49, plotA50_84, ncol=3, nrow=2, top = grid::textGrob('Breast cancer in females by morphology and ER status', gp=grid::gpar(fontsize=8)))





```

Figure 1 shows the trends in the incidence of Infiltrating Ductal Cancer (IDC) and Ductal Carcinoma In Situ (DCIS) breast cancers relative to 1980. A) depicts overall trends; B) depicts trends in younger women; and C) depicts trends in older women. The dashed line corresponds to a value of 1.
