---
toc: true
title: "Simple Imputation -  malignant breast cancer by ER using SEER 12, 1992–2020"
author: "AMF"
format: html
editor: visual
date: "January 31, 2024"
date-modified: last-modified
---



#### Data Source
National Cancer Institute. Surveillance, Epidemiology, and End Results (SEER) Program Populations (1975-2020). Available at: [www.seer.cancer.gov/popdata](https://www.seer.cancer.gov/popdata). Released May 2023. National Cancer Institute, DCCPS, Surveillance Research Program, February 2022S ed.

## Morphology
#### Infiltrating Ductal Carcinoma (IDC)
:::{.callout-note}
-   **8500/3**: Infiltrating duct carcinoma, NOS'
-   **8523/3**: Infiltrating duct mixed with other types of carcinoma'
:::

ER Status Recode for Breast Cancer (1990+): 'Positive', 'Negative', 'Borderline/Unknown', 'Recode not available


#### load required libraries
```{r, message=FALSE, warning=FALSE}
# Load necessary library for data fetching
library(readr)
library(ggplot2)
library(dplyr)
library(gridExtra)

```

##### Importing Encrypted SEER12 Data Files from GitHub Repository

Discuss the process involved in accessing locked data files from SEER12 that have been stored on GitHub.

```{r, message=FALSE, warning=FALSE}
# Read the dataset from the provided URL
url <- "https://raw.githubusercontent.com/filhoalm/Breast_cancer/main/Incidence/see12/seer12_er.csv"
df <- read_csv(url)

```

```{r, message=FALSE, warning=FALSE, echo=FALSE}


#| label: fig1
#| fig-cap: "Figure 1. Trends in the prevalence of estrogen receptor (ER) status for invasive female breast cancer cases in the United States are depicted in two parts. Figure A presents the annual proportions of breast cancer cases with unspecified ER status. Figure B details the yearly proportions of breast cancers with confirmed ER status that are ER-negative."
#| 
names(df)<-c("age", "year", "er", "rate", "cases", "py")
#unique(df$er)
df$er[df$er==3]<-2 

# Aggregate cases by age, year, and er
df_a <- df %>%
  group_by(age, year, er) %>%
  summarise(total_cases = sum(cases), .groups = 'drop')

# Recode the race and stage values
df_a$er <- factor(df_a$er, levels = c(0, 1, 2),
                labels = c("Positive", "Negative", "Unknown"))

# Filter out rows with "Year of diagnosis" in the 'year' column and convert to numeric
#df <- subset(df, df$year != "Year of diagnosis")
df_a$year <- as.numeric(df_a$year) + 1991
#unique(df_a$year)
df1<-df_a
df2<-subset(df1, df1$age >=40)


# Sum the number of ER cases by year and age
df_summary <- df2 %>%
  group_by(year, age) %>%
  summarise(total_cases = sum(total_cases), .groups = 'drop') # Sum of 'cases' within each 'year' and 'age' group
df3<-merge(df_summary, df2, by=c("age", "year"))
# Sum the number of ER cases by year and age
df_s <- df2 %>%
  group_by(year, er) %>%
  summarise(total_cases = sum(total_cases), .groups = 'drop') # Sum of 'cases' within each 'year' and 'age' group
# Sum the number of ER cases by year and age
df_s1 <- df2 %>%
  group_by(year) %>%
  summarise(total_cases = sum(total_cases), .groups = 'drop') # Sum of 'cases' within each 'year' and 'age' group
df4<-merge(df_s, df_s1, by=c("year"))
# Proportion of unknown
df4$prop<-(df4$total_cases.x / df4$total_cases.y)
# Filter for 'Unknown' ER only
df_unknown_er <- df4 %>%
  filter(er == "Unknown")

df_unknown_er<-subset(df_unknown_er, df_unknown_er$year>=1992)

# Create a line plot in Lance style
a<-ggplot(df_unknown_er, aes(x = year, y = prop, group = 1)) +
  geom_line(color = "black", size = 1) +  # Change line color and size as needed
  ylim(0,0.5)+
  geom_point(color = "black", size = 2) + # Add points and customize their color and size as needed
  theme_minimal() + # Lance style typically implies a minimalistic theme
  labs(
    title = "Year at diagnosis",
    x = "Year",
    y = "% of Unknown ER"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5), # Center the plot title
    text = element_text(size = 12) # Change text size as needed
  )


## Proportion Age at diagnosis

df2a<-subset(df2, df2$year>=1992 & df2$year <=2008) #1991 represents 1992-2020

# Sum the number of ER cases by year and age
dfb <- df2a %>%
  group_by(age) %>%
  summarise(total_cases = sum(total_cases), .groups = 'drop') # Sum of 'cases' within each 'year' and 'age' group

# Sum the number of ER cases by year and age
dfc <- df2a %>%
  group_by(age, er) %>%
  summarise(total_cases = sum(total_cases), .groups = 'drop') # Sum of 'cases' within each 'year' and 'age' group
df4a<-merge(dfc, dfb, by=c("age"))

# Proportion of unknown
df4a$prop<-(df4a$total_cases.x / df4a$total_cases.y)

# Filter for 'Unknown' ER only
df_age <- df4a %>%
  filter(er == "Negative")

# Create a line plot in Lance style
b<-ggplot(df_age, aes(x = age, y = prop, group = 1)) +
  geom_line(color = "black", size = 1) +  # Change line color and size as needed
  ylim(0,0.5)+
  geom_point(color = "black", size = 2) + # Add points and customize their color and size as needed
  theme_minimal() + # Lance style typically implies a minimalistic theme
  labs(
    title = "Age at diagnosis",
    x = "Age",
    y = "% ER Negative"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5), # Center the plot title
    text = element_text(size = 12) # Change text size as needed
  )




grid.arrange(a,b,
             ncol=2,
             top = grid::textGrob('', gp=grid::gpar(fontsize=5)))


```







##### Simple imputation for malignant breast cancer by ER

A simple imputation method to correct the rates of ER-positive and ER-negative was based on the method proposed by[Anderson at al., 2011](https://pubmed.ncbi.nlm.nih.gov/21753181/)
