---
author: "Notes from: IJ; AMF; GG; PR"
title: "SEER 8 - Breast cancer: invasive vs in situ"
date: "Today's Date"
output: pdf_document
---

## Data source

Institute NC. Surveillance, Epidemiology, and End Results (SEER 9,) Program Populations (1975-2018). (www.seer.cancer.gov/popdata), National Cancer Institute, DCCPS, Surveillance Research Program, released May 2023. February 2022S ed.: National Cancer Institute, 2022.

#### Morphology

###### DCIS

'8201/2: Cribriform carcinoma in situ','8500/2: Intraductal carcinoma, noninfiltrating, NOS','8501/2: Comedocarcinoma, noninfiltrating','8503/2: Noninfiltrating intraductal papillary adenocarcinoma','8507/2: Intraductal micropapillary carcinoma','8523/2: Intraductal with other types of carcinoma in situ' AND {Extent of Disease.ER Status Recode Breast Cancer (1990+)} = 'Positive','Negative','Borderline/Unknown','Recode not available'

###### IDC

'8500/3: Infiltrating duct carcinoma, NOS','8523/3: Infiltrating duct mixed with other types of carcinoma' AND {Extent of Disease.ER Status Recode Breast Cancer (1990+)} = 'Positive','Negative','Borderline/Unknown','Recode not available'



##### Check the data extracted from SEER 8.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(Rcan)
library(ggplot2)
```

```{r load_data}
df1 <- read.csv("https://raw.githubusercontent.com/filhoalm/Breast_cancer/main/dataCheck/dcis.csv")
names(df1) <- c("year", "age", "er", "asr", "count", "py", "std")
df1$cancer <- "DCIS"

df2 <- read.csv("https://raw.githubusercontent.com/filhoalm/Breast_cancer/main/dataCheck/idc.csv")
names(df2) <- c("year", "age", "er", "asr", "count", "py", "std")
df2$cancer <- "IDC"

```


```{r load_data2}
df <- rbind(df1, df2)
df$year <- df$year + 1974

df <- subset(df, age >= 30 & age <= 84 & year >= 1975)

df$er_label <- ifelse(df$er == 0, "positive",
                      ifelse(df$er == 1, "negative",
                             ifelse(df$er == 2, "bord/unkw", "not available")))

df$age_group <- cut(df$age, breaks = seq(29, 85, 5), labels = paste(seq(30, 80, 5), "-", seq(34, 84, 5)))

df$age_index <- as.integer(df$age_group) + 6

df4 <- df[c(1, 5, 6, 8, 9, 11)]

result <- csu_asr(df4, "age_index", "count", "py",
                  group_by = c("cancer", "er_label", "year"),
                  var_age_group = c("er_label", "year"))

df5 <- subset(result, year >= 1992 )

```


## ASR have been computed for the age group 30-84

```{r create_plot}
ggplot(df5, aes(x = year, y = asr, color = er_label, shape = cancer)) +
  geom_point() +
  geom_line() +
  labs(x = "Year", y = "ASR", color = "ER Label", shape = "Cancer") +
  theme_minimal()
```

```{r load_data3}
df <- rbind(df1, df2)
df$year <- df$year + 1974

df <- subset(df, age >= 30 & age <= 84 & year >= 1975)

df$er_label <- ifelse(df$er == 0, "positive",
                      ifelse(df$er == 1, "negative",
                             ifelse(df$er == 2, "bord/unkw", "not available")))

df$age_group <- cut(df$age, breaks = seq(29, 85, 5), labels = paste(seq(30, 80, 5), "-", seq(34, 84, 5)))

df$age_index <- as.integer(df$age_group) + 6

dfa<-subset(df, df$age_index >= 7 & df$age_index <=10)

df4 <- dfa[c(1, 5, 6, 8, 9, 11)]

result1 <- csu_asr(df4, "age_index", "count", "py",
                  group_by = c("cancer", "er_label", "year"),
                  var_age_group = c("er_label", "year"))

df6 <- subset(result1, year >= 1992 )

```
## ASR have been computed for the age group 30-49

```{r create_plot2}
ggplot(df6, aes(x = year, y = asr, color = er_label, shape = cancer)) +
  geom_point() +
  geom_line() +
  labs(x = "Year", y = "ASR", color = "ER Label", shape = "Cancer") +
  theme_minimal()

```

```{r load_data4}
df <- rbind(df1, df2)
df$year <- df$year + 1974

df <- subset(df, age >= 30 & age <= 84 & year >= 1975)

df$er_label <- ifelse(df$er == 0, "positive",
                      ifelse(df$er == 1, "negative",
                             ifelse(df$er == 2, "bord/unkw", "not available")))

df$age_group <- cut(df$age, breaks = seq(29, 85, 5), labels = paste(seq(30, 80, 5), "-", seq(34, 84, 5)))

df$age_index <- as.integer(df$age_group) + 6

dfb<-subset(df, df$age_index >= 11)

df4 <- dfb[c(1, 5, 6, 8, 9, 11)]

result2 <- csu_asr(df4, "age_index", "count", "py",
                  group_by = c("cancer", "er_label", "year"),
                  var_age_group = c("er_label", "year"))

df7 <- subset(result2, year >= 1992 )

```
## ASR have been computed for the age group 50-84+

```{r create_plot3}
ggplot(df7, aes(x = year, y = asr, color = er_label, shape = cancer)) +
  geom_point() +
  geom_line() +
  labs(x = "Year", y = "ASR", color = "ER Label", shape = "Cancer") +
  theme_minimal()

```